The Calculus of Linear Constructions

### 2022-01-17 writing/draft7/draft7.tex
draft

\subsection{Meta Theory} \label{meta}
In this section, we focus our discussion on the properties of CLC. First, we show the type soundness of CLC through the \textit{subject reduction} theorem. Next, we show that CLC is an effective linear type theory through the \textit{linearity} theorem. Furthermore, we show that the \textit{promotion} and \textit{dereliction} rules can be encoded as $\eta$-expansions. Finally, we construct a reduction preserving erasure function that maps well-typed CLC terms to well-typed CC$\omega$ terms, showing that CLC is strongly normalizing.

All proofs have been formalized in Coq with help from the Autosubst \cite{autosubst} library. The Coq development is publicly available on the first author's Github repository. To the best of our knowledge, this is the first machined checked formalization of a linear dependently type theory. We give a hand written version of the proof in an accompanying technical report.

@@

The proof of confluence is entirely standard using parallel step technique \cite{takahashi}. The presence of linear types does not pose any complications as reductions are untyped.

@@

An indexing function can now be constructed as a wrapper for $\texttt{nth}$ as shown in Figure \ref{index}. Given an array $\texttt{Array}\ A\ n$, an offset $m$ and proof that $m < n$, $\texttt{nth}$ can be used to extract the capability at offset $m$, allowing $\get$ to retrieve the stored value. Finally, the array proof is restored and returned alongside the indexing result. This definition of $\texttt{index}$ is safe because the proof $m < n$ ensures that offsets used for indexing are always within bounds.

\begin{figure}[h]
  \caption{Array Indexing}
  \begin{minted}[escapeinside=??,mathescape=true]{coq}
Definition index
  (A : U)
  (m n : ?$\mathbb{N}$?)
  (pf : m < n)
  (a : Array A n) 
: ?$\SigmaL$? _ : A. Array A n
  \end{minted}
  \label{index}
  \Description{}
\end{figure}

@@

\begin{figure}[h]
  \caption{Inductive Session Type}
  \begin{minted}{coq}
Inductive session : U :=
| SEND : U -> session -> session
| RECV : U -> session -> session
| END : session.
  \end{minted}
  \label{session}
  \Description{}
\end{figure}

@@

\begin{figure}[h]
  \caption{Linear Lists}
  \begin{minted}{coq}
Inductive list (A : U) : L :=
| nil : list A
| cons : A -> list A -> list A.

Fixpoint append (A : U) : list A -o 
  list A -o (list A -o list A) -o list A 
:=
  fun ls1 ls2 k =>
    match ls1 with
    | nil => k ls2
    | (cons h t) as _cons_ =>
      append _ t ls2 
        (fun res => k (_cons_ h res))
    end.
  \end{minted}
  \label{list}
  \Description{}
\end{figure}

@@

\begin{lemma}\label{convapp}
  For terms $m_1, m_2, n_1, n_2$, if there exists equalities $m_1 \equiv m_2$ and $n_1 \equiv n_2$, then there exists equality $(m_1\ n_1) \equiv (m_2\ n_2)$.
\end{lemma}
\begin{proof}
  By transitivity of $\equiv$ and applying rules \rname{Join}, \rname{Step-AppL}, \rname{Step-AppR}.
\end{proof}

\begin{lemma}\label{convlam}
  For terms $A_1, A_2, m_1, m_2$ and sort $s$, if there exists equalities $A_1 \equiv A_2$ and $m_1 \equiv m_2$, then there exists equality $\lambda x \stype{s} A_1.m_1 \equiv \lambda x \stype{s} A_2.m_2$.
\end{lemma}
\begin{proof}
  By transitivity of $\equiv$ and applying rules \rname{Join}, \rname{Step-$\lambda$L}, \rname{Step-$\lambda$R}.
\end{proof}

\begin{lemma}\label{convarrow}
  For terms $A_1, A_2, B_1, B_2$ and sort $s$, if there exists equalities $A_1 \equiv A_2$ and $B_1 \equiv B_2$, then there exists equality $(x \stype{s} A_1) \rightarrow B_1 \equiv (x \stype{s} A_2) \rightarrow B_2$.
\end{lemma}
\begin{proof}
  By transitivity of $\equiv$ and applying rules \rname{Join}, \rname{Step-L$\rightarrow$}, \rname{Step-R$\rightarrow$}.
\end{proof}

\begin{lemma}\label{convlolli}
  For terms $A_1, A_2, B_1, B_2$ and sort $s$, if there exists equalities $A_1 \equiv A_2$ and $B_1 \equiv B_2$, then there exists equality $(x \stype{s} A_1) \multimap B_1 \equiv (x \stype{s} A_2) \multimap B_2$.
\end{lemma}
\begin{proof}
  By transitivity of $\equiv$ and applying rules \rname{Join}, \rname{Step-L$\multimap$}, \rname{Step-R$\multimap$}.
\end{proof}

@@

\begin{lemma}\label{pstepcompat}
  For terms $m, n$ and maps $\sigma, \tau$ from variables to terms, if there exists parallel reduction $m \pstep n$ and parallel map reduction $\sigma \pstep \tau$, there exists parallel reduction $m[\sigma] \pstep n[\tau]$.
\end{lemma}
\begin{proof}
  By induction on the derivation of $m \pstep n$.
\end{proof}

@@

\begin{lemma}\label{strip}
  Strip lemma. For terms $m, m_1, m_2$, if there is parallel reduction $m \pstep m_1$ and reduction $m \red m_2$, then there exists term $m'$ such that $m_1 \red m'$ and $m_2 \pstep m'$.
\end{lemma}
\begin{proof}
  By induction on the derivation of $m \pstep m_1$, utilizing transitivity of $\red$ and Lemmas \ref{steppstep}, \ref{pstepred}, \ref{diamond}.
\end{proof}

@@

\begin{corollary}\label{redarrowinv}
  For terms $A, B, m$ and sort $s$, if there is reduction $(x \stype{s} A) \rightarrow B \red m$, then there exists $A', B'$ such that there are reductions $A \red A'$, $B \red B'$ and $m = (x \stype{s} A') \rightarrow B'$.
\end{corollary}

\begin{corollary}\label{redlolliinv}
  For terms $A, B, m$ and sort $s$, if there is reduction $(x \stype{s} A) \multimap B \red m$, then there exists $A', B'$ such that there are reductions $A \red A'$, $B \red B'$ and $m = (x \stype{s} A') \multimap B'$.
\end{corollary}

\begin{corollary}\label{redlaminv}
  For terms $A, m, n$ and sort $s$, if there is reduction $\lambda x \stype{s} A.m \red n$, then there exists $A',m'$ such that there are reductions $A \red A'$, $m \red m'$ and $n = \lambda x \stype{s} A'.m'$.
\end{corollary}

\begin{corollary}\label{sortinj}
  For sorts $s, t$ and levels $i, j$, if there is equality $s_i \equiv t_j$, then there is $s = t$ and $i = j$.
\end{corollary}

@@

\begin{lemma}\label{mergepureinv}
  For contexts $\Gamma_1, \Gamma_2, \Gamma$, if there is $\mrg{\Gamma_1}{\Gamma_2}{\Gamma}$ and $\pure{\Gamma}$, then there is $\pure{\Gamma_1}$ and $\pure{\Gamma_2}$.
\end{lemma}
\begin{proof}
  By induction on the derivation of $\mrg{\Gamma_1}{\Gamma_2}{\Gamma}$.
\end{proof}

\begin{lemma}\label{mergepure1}
  For contexts $\Gamma_1, \Gamma_2, \Gamma$ , if there is $\mrg{\Gamma_1}{\Gamma_2}{\Gamma}$ and $\pure{\Gamma_1}$, then there is $\Gamma = \Gamma_2$.
\end{lemma}
\begin{proof}
  By induction on the derivation of $\mrg{\Gamma_1}{\Gamma_2}{\Gamma}$.
\end{proof}

\begin{lemma}\label{mergepure2}
  For contexts $\Gamma_1, \Gamma_2, \Gamma$ , if there is $\mrg{\Gamma_1}{\Gamma_2}{\Gamma}$ and $\pure{\Gamma_2}$, then there is $\Gamma = \Gamma_1$.
\end{lemma}
\begin{proof}
  By induction on the derivation of $\mrg{\Gamma_1}{\Gamma_2}{\Gamma}$.
\end{proof}

\begin{lemma}\label{mergepurepure}
  For contexts $\Gamma_1, \Gamma_2, \Gamma$, if there is $\mrg{\Gamma_1}{\Gamma_2}{\Gamma}$, and also $\pure{\Gamma_1}$, $\pure{\Gamma_2}$, then there is $\pure{\Gamma}$.
\end{lemma}
\begin{proof}
  By induction on the derivation of $\mrg{\Gamma_1}{\Gamma_2}{\Gamma}$.
\end{proof}

\begin{lemma}\label{mergepureeq}
  For contexts $\Gamma_1, \Gamma_2, \Gamma$, if there is $\mrg{\Gamma_1}{\Gamma_2}{\Gamma}$, and also $\pure{\Gamma_1}$, $\pure{\Gamma_2}$, then there is $\Gamma_1 = \Gamma_2$.
\end{lemma}
\begin{proof}
  By induction on the derivation of $\mrg{\Gamma_1}{\Gamma_2}{\Gamma}$.
\end{proof}

\begin{lemma}\label{mergerere}
  For contexts $\Gamma_1, \Gamma_2, \Gamma$, if there is $\mrg{\Gamma_1}{\Gamma_2}{\Gamma}$, then there is $\overline{\Gamma_1} = \overline{\Gamma}$ and $\overline{\Gamma_2} = \overline{\Gamma}$.
\end{lemma}
\begin{proof}
  By induction on the derivation of $\mrg{\Gamma_1}{\Gamma_2}{\Gamma}$.
\end{proof}

@@

\begin{lemma}\label{mergesplit1}
  For contexts $\Gamma_1, \Gamma_2, \Gamma, \Delta_1, \Delta_2$, if there is $\mrg{\Gamma_1}{\Gamma_2}{\Gamma}$ and $\mrg{\Delta_1}{\Delta_2}{\Gamma_1}$, then there exists $\Delta$ such that $\mrg{\Delta_1}{\Gamma_2}{\Delta}$ and $\mrg{\Delta}{\Delta_2}{\Gamma}$.
\end{lemma}
\begin{proof}
  By induction on the derivation of $\mrg{\Gamma_1}{\Gamma_2}{\Gamma}$.
\end{proof}

\begin{lemma}
  For contexts $\Gamma_1, \Gamma_2, \Gamma, \Delta_1, \Delta_2$, if there is $\mrg{\Gamma_1}{\Gamma_2}{\Gamma}$ and $\mrg{\Delta_1}{\Delta_2}{\Gamma_1}$, then there exists $\Delta$ such that $\mrg{\Delta_2}{\Gamma_2}{\Delta}$ and $\mrg{\Delta_1}{\Delta}{\Gamma}$.
\end{lemma}
\begin{proof}
  By induction on the derivation of $\mrg{\Gamma_1}{\Gamma_2}{\Gamma}$.
\end{proof}

@@

\begin{lemma}\label{sub1conv}
  For terms $A, B, C$, if there is $A \prec B$ and $B \equiv C$, then there is $A \preceq C$.
\end{lemma}
\begin{proof}
  By \rname{$\prec$-$\preceq$} and the transitivity of equality $\equiv$.
\end{proof}

\begin{lemma}\label{convsub1}
  For terms $A, B, C$, if there is $A \equiv B$ and $B \prec C$, then there is $A \preceq C$.
\end{lemma}
\begin{proof}
  By \rname{$\prec$-$\preceq$} and the transitivity of equality $\equiv$.
\end{proof}

@@

\begin{lemma}\label{sub1trans}
  For terms $A, B, C, D$, if there is $A \prec B$, $B \equiv C$ and $C \prec D$, then there is $A \preceq D$.
\end{lemma}
\begin{proof}
  By induction on the derivation of $A \prec B$, definition of $\prec$ and Lemmas \ref{sub1sub}, \ref{sub1conv}, \ref{convsub1}.
\end{proof}

\begin{lemma}
  For terms $A, B, C$, if there is $A \preceq B$ and $B \preceq C$, then there is $A \preceq C$.
\end{lemma}
\begin{proof}
  By transitivity of $\equiv$, rule \rname{$\prec$-$\preceq$} and Lemma \ref{sub1trans}.
\end{proof}

\begin{lemma}
  For sorts $s, t$ and natural numbers $i, j$, if there is $s_i \preceq t_j$, then there is $s = t$ and $i \leq j$.
\end{lemma}
\begin{proof}
  By transitivity of $\equiv$ and Corollary \ref{sortinj}.
\end{proof}

\begin{lemma}
  For terms $A_1, A_2, B_1, B_2$ and sorts $s, t$, if there is $(x \stype{s} A_1) \rightarrow B_1 \preceq (x \stype{t} A_2) \rightarrow B_2$, then there is $A_1 \equiv A_2$ and $B_1 \preceq B_2$ and $s = t$.
\end{lemma}
\begin{proof}
  By transitivity of $\equiv$ and Corollary \ref{arrowinj}.
\end{proof}

\begin{lemma}
  For terms $A_1, A_2, B_1, B_2$ and sorts $s, t$, if there is $(x \stype{s} A_1) \multimap B_1 \preceq (x \stype{t} A_2) \multimap B_2$, then there is $A_1 \equiv A_2$ and $B_1 \preceq B_2$ and $s = t$.
\end{lemma}
\begin{proof}
  By transitivity of $\equiv$ and Corollary \ref{lolliinj}.
\end{proof}

@@

\begin{lemma}\label{uarrowinv}
  For any context $\Gamma$ and terms $A, B, s$, if there is $\Gamma \vdash (x \utype A) \rightarrow B : s$, then there exists sort $t$ and natural number $i$ such that $\Gamma \vdash A : U_i$ and $\Gamma, x \utype A \vdash B : t_i$.
\end{lemma}
\begin{proof}
  By induction on the derivation of $\Gamma \vdash (x \utype A) \rightarrow B : s$.
\end{proof}

\begin{lemma}\label{larrowinv}
  For any context $\Gamma$ and terms $A, B, s$, if there is $\Gamma \vdash (x \ltype A)\rightarrow B : s$, then there exists sort $t$ and natural number $i$ such that $\Gamma \vdash A : L_i$ and $\Gamma \vdash B : t_i$.
\end{lemma}
\begin{proof}
  By induction on the derivation of $\Gamma \vdash (x \ltype A) \rightarrow B : s$.
\end{proof}

\begin{lemma}\label{ulolliinv}
  For any context $\Gamma$ and terms $A, B, s$, if there is $\Gamma \vdash (x \utype A) \multimap B : s$, then there exists sort $t$ and natural number $i$ such that $\Gamma \vdash A : U_i$ and $\Gamma, x \utype A \vdash B : t_i$.
\end{lemma}
\begin{proof}
  By induction on the derivation of $\Gamma \vdash (x \utype A) \multimap B : s$.
\end{proof}

\begin{lemma}\label{llolliinv}
  For any context $\Gamma$ and terms $A, B, s$, if there is $\Gamma \vdash (x \ltype A)\multimap B : s$, then there exists sort $t$ and natural number $i$ such that $\Gamma \vdash A : L_i$ and $\Gamma \vdash B : t_i$.
\end{lemma}
\begin{proof}
  By induction on the derivation of $\Gamma \vdash (x \ltype A) \multimap B : s$.
\end{proof}

@@

\begin{lemma}\label{arrowlaminv}
  For any context $\Gamma$, terms $A, A', B, n$, sorts $s, s', t$ and natural number $i$, if there is $\overline{\Gamma} \vdash (x \stype{s'} A') \rightarrow B : t_i$ and $\Gamma \vdash \lambda x \stype{s} A . n : (x \stype{s'} A') \rightarrow B$, then there is $\Gamma, x \stype{s'} A' \vdash n : B$.
\end{lemma}
\begin{proof}
  Direct consequence of Lemmas \ref{uarrowinv}, \ref{larrowinv} and \ref{arrowlaminvx}.
\end{proof}

\begin{lemma}\label{lollilaminv}
  For any context $\Gamma$, terms $A, A', B, n$, sorts $s, s', t$ and natural number $i$, if there is $\overline{\Gamma} \vdash (x \stype{s'} A') \multimap B : t_i$ and $\Gamma \vdash \lambda x \stype{s} A . n : (x \stype{s'} A') \multimap B$, then there is $\Gamma, x \stype{s'} A' \vdash n : B$.
\end{lemma}
\begin{proof}
  Direct consequence of Lemmas \ref{ulolliinv}, \ref{llolliinv} and \ref{lollilaminvx}.
\end{proof}

@@

\begin{lemma}\label{agreerenpure}
  For contexts $\Gamma, \Gamma'$ and mapping $\xi$, if there is $agreeR\ \xi\ \Gamma\ \Gamma'$ and $\pure{\Gamma}$, then there is $\pure{\Gamma'}$.
\end{lemma}
\begin{proof}
  By induction on the derivation of $agreeR\ \xi\ \Gamma\ \Gamma'$.
\end{proof}

@@

\begin{lemma}\label{mergeagreereninv}
  For contexts $\Gamma, \Gamma', \Gamma_1, \Gamma_2$ and mapping $\xi$, if there is $agreeR\ \xi\ \Gamma\ \Gamma'$ and $\mrg{\Gamma_1}{\Gamma_2}{\Gamma}$, then there exists $\Gamma_1', \Gamma_2'$ such that $\mrg{\Gamma_1'}{\Gamma_2'}{\Gamma'}$, and $agreeR\ \xi\ \Gamma_1\ \Gamma_1'$ and $agreeR\ \xi\ \Gamma_2\ \Gamma_2'$.
\end{lemma}
\begin{proof}
  By induction on the derivation of $agreeR\ \xi\ \Gamma\ \Gamma'$ and lemmas in Section \ref{agreeRprop}.
\end{proof}

>>>>

\subsection{Meta Theory} \label{meta}
In this section, we focus our discussion on the properties of CLC. First, we show the type soundness of CLC through the \textit{subject reduction} theorem. Next, through the \textit{linearity} theorem we illustrate how CLC can effectively control usage of resource. Furthermore, we demonstrate that the \textit{promotion} and \textit{dereliction} rules can be encoded as $\eta$-expansions. Finally, we construct a reduction preserving erasure function that maps well-typed CLC terms to well-typed CC$\omega$ terms, proving that CLC is strongly normalizing.

All proofs have been formalized in Coq with help from the Autosubst \cite{autosubst} library. The Coq development is publicly available on the author's Github repository. To the best of our knowledge, this is the first machine checked formalization of a linear dependently type theory. We give a hand written version of the proof in an accompanying technical report.

@@

The proof of confluence is entirely standard using parallel reduction technique \cite{takahashi}. The presence of linear types does not pose any complications as reductions are untyped.

@@

An indexing function can now be constructed as a wrapper for $\texttt{nth}$ with its signature shown in Figure \ref{index}. Given an array $\texttt{Array}\ A\ n$, an offset $m$ and proof that $m < n$, $\texttt{nth}$ can be used to extract the capability at offset $m$, allowing $\get$ to retrieve the stored value. Finally, the array proof is restored and returned alongside the indexing result. This definition of $\texttt{index}$ is safe because the proof $m < n$ ensures that offsets used for indexing are always within bounds.

\begin{figure}[h]
  \caption{Array Indexing}
  \begin{minted}[escapeinside=??,mathescape=true]{coq}
Definition index
  (A : U)
  (m n : ?$\mathbb{N}$?)
  (pf : m < n)
  (a : Array A n) 
: ?$\SigmaL$? _ : A. Array A n.
  \end{minted}
  \label{index}
  \Description{}
\end{figure}

@@

\begin{figure}[h]
  \caption{Inductive Session Type}
  \begin{minted}[escapeinside=??,mathescape=true]{coq}
Inductive session : U :=
| SEND : U ?$\rightarrow$? session ?$\rightarrow$? session
| RECV : U ?$\rightarrow$? session ?$\rightarrow$? session
| END : session.
  \end{minted}
  \label{session}
  \Description{}
\end{figure}

@@

\begin{figure}[h]
  \caption{Linear Lists}
  \begin{minted}[escapeinside=??,mathescape=true]{coq}
Inductive list (A : U) : L :=
| nil : list A
| cons : A ?$\rightarrow$? list A ?$\rightarrow$? list A.

Fixpoint append (A : U) : list A ?$\multimap$? 
  list A ?$\multimap$? (list A ?$\multimap$? list A) ?$\multimap$? list A 
:=
  fun ls1 ls2 k ?$\Rightarrow$?
    match ls1 with
    | nil ?$\Rightarrow$? k ls2
    | (cons h t) as _cons_ ?$\Rightarrow$?
      append _ t ls2 
        (fun res ?$\Rightarrow$? k (_cons_ h res))
    end.
  \end{minted}
  \label{list}
  \Description{}
\end{figure}

@@

\begin{lemma}\label{convapp}
  For terms $m_1, m_2, n_1, n_2$, if there exist equalities $m_1 \equiv m_2$ and $n_1 \equiv n_2$, then there exists equality $(m_1\ n_1) \equiv (m_2\ n_2)$.
\end{lemma}
\begin{proof}
  By transitivity of $\equiv$ and applying rules \rname{Join}, \rname{Step-AppL}, \rname{Step-AppR}.
\end{proof}

\begin{lemma}\label{convlam}
  For terms $A_1, A_2, m_1, m_2$ and sort $s$, if there exist equalities $A_1 \equiv A_2$ and $m_1 \equiv m_2$, then there exists equality $\lambda x \stype{s} A_1.m_1 \equiv \lambda x \stype{s} A_2.m_2$.
\end{lemma}
\begin{proof}
  By transitivity of $\equiv$ and applying rules \rname{Join}, \rname{Step-$\lambda$L}, \rname{Step-$\lambda$R}.
\end{proof}

\begin{lemma}\label{convarrow}
  For terms $A_1, A_2, B_1, B_2$ and sort $s$, if there exist equalities $A_1 \equiv A_2$ and $B_1 \equiv B_2$, then there exists equality $(x \stype{s} A_1) \rightarrow B_1 \equiv (x \stype{s} A_2) \rightarrow B_2$.
\end{lemma}
\begin{proof}
  By transitivity of $\equiv$ and applying rules \rname{Join}, \rname{Step-L$\rightarrow$}, \rname{Step-R$\rightarrow$}.
\end{proof}

\begin{lemma}\label{convlolli}
  For terms $A_1, A_2, B_1, B_2$ and sort $s$, if there exist equalities $A_1 \equiv A_2$ and $B_1 \equiv B_2$, then there exists equality $(x \stype{s} A_1) \multimap B_1 \equiv (x \stype{s} A_2) \multimap B_2$.
\end{lemma}
\begin{proof}
  By transitivity of $\equiv$ and applying rules \rname{Join}, \rname{Step-L$\multimap$}, \rname{Step-R$\multimap$}.
\end{proof}

@@

\begin{lemma}\label{pstepcompat}
  For terms $m, n$ and maps $\sigma, \tau$ from variables to terms, if there exist parallel reduction $m \pstep n$ and parallel map reduction $\sigma \pstep \tau$, there exists parallel reduction $m[\sigma] \pstep n[\tau]$.
\end{lemma}
\begin{proof}
  By induction on the derivation of $m \pstep n$.
\end{proof}

@@

\begin{lemma}\label{strip}
  Strip lemma. For terms $m, m_1, m_2$, if there are parallel reductions $m \pstep m_1$ and $m \red m_2$, then there exists term $m'$ such that $m_1 \red m'$ and $m_2 \pstep m'$.
\end{lemma}
\begin{proof}
  By induction on the derivation of $m \pstep m_1$, utilizing transitivity of $\red$ and Lemmas \ref{steppstep}, \ref{pstepred}, \ref{diamond}.
\end{proof}

@@

\begin{corollary}\label{redarrowinv}
  For terms $A, B, m$ and sort $s$, if there is reduction $(x \stype{s} A) \rightarrow B \red m$, then there exist $A'$ and $B'$ such that there are reductions $A \red A'$, $B \red B'$ and $m = (x \stype{s} A') \rightarrow B'$.
\end{corollary}

\begin{corollary}\label{redlolliinv}
  For terms $A, B, m$ and sort $s$, if there is reduction $(x \stype{s} A) \multimap B \red m$, then there exist $A'$ and $B'$ such that there are reductions $A \red A'$, $B \red B'$ and $m = (x \stype{s} A') \multimap B'$.
\end{corollary}

\begin{corollary}\label{redlaminv}
  For terms $A, m, n$ and sort $s$, if there is reduction $\lambda x \stype{s} A.m \red n$, then there exist $A'$ and $m'$ such that there are reductions $A \red A'$, $m \red m'$ and $n = \lambda x \stype{s} A'.m'$.
\end{corollary}

\begin{corollary}\label{sortinj}
  For sorts $s, t$ and levels $i, j$, if there is equality $s_i \equiv t_j$, then there are $s = t$ and $i = j$.
\end{corollary}

@@

\begin{lemma}\label{mergepureinv}
  For contexts $\Gamma_1, \Gamma_2, \Gamma$, if there are $\mrg{\Gamma_1}{\Gamma_2}{\Gamma}$ and $\pure{\Gamma}$, then there are $\pure{\Gamma_1}$ and $\pure{\Gamma_2}$.
\end{lemma}
\begin{proof}
  By induction on the derivation of $\mrg{\Gamma_1}{\Gamma_2}{\Gamma}$.
\end{proof}

\begin{lemma}\label{mergepure1}
  For contexts $\Gamma_1, \Gamma_2, \Gamma$ , if there are $\mrg{\Gamma_1}{\Gamma_2}{\Gamma}$ and $\pure{\Gamma_1}$, then there is $\Gamma = \Gamma_2$.
\end{lemma}
\begin{proof}
  By induction on the derivation of $\mrg{\Gamma_1}{\Gamma_2}{\Gamma}$.
\end{proof}

\begin{lemma}\label{mergepure2}
  For contexts $\Gamma_1, \Gamma_2, \Gamma$ , if there are $\mrg{\Gamma_1}{\Gamma_2}{\Gamma}$ and $\pure{\Gamma_2}$, then there is $\Gamma = \Gamma_1$.
\end{lemma}
\begin{proof}
  By induction on the derivation of $\mrg{\Gamma_1}{\Gamma_2}{\Gamma}$.
\end{proof}

\begin{lemma}\label{mergepurepure}
  For contexts $\Gamma_1, \Gamma_2, \Gamma$, if there are $\mrg{\Gamma_1}{\Gamma_2}{\Gamma}$, $\pure{\Gamma_1}$ and $\pure{\Gamma_2}$, then there is $\pure{\Gamma}$.
\end{lemma}
\begin{proof}
  By induction on the derivation of $\mrg{\Gamma_1}{\Gamma_2}{\Gamma}$.
\end{proof}

\begin{lemma}\label{mergepureeq}
  For contexts $\Gamma_1, \Gamma_2, \Gamma$, if there are $\mrg{\Gamma_1}{\Gamma_2}{\Gamma}$, $\pure{\Gamma_1}$ and $\pure{\Gamma_2}$, then there is $\Gamma_1 = \Gamma_2$.
\end{lemma}
\begin{proof}
  By induction on the derivation of $\mrg{\Gamma_1}{\Gamma_2}{\Gamma}$.
\end{proof}

\begin{lemma}\label{mergerere}
  For contexts $\Gamma_1, \Gamma_2, \Gamma$, if there is $\mrg{\Gamma_1}{\Gamma_2}{\Gamma}$, then there are $\overline{\Gamma_1} = \overline{\Gamma}$ and $\overline{\Gamma_2} = \overline{\Gamma}$.
\end{lemma}
\begin{proof}
  By induction on the derivation of $\mrg{\Gamma_1}{\Gamma_2}{\Gamma}$.
\end{proof}

@@

\begin{lemma}\label{mergesplit1}
  For contexts $\Gamma_1, \Gamma_2, \Gamma, \Delta_1, \Delta_2$, if $\mrg{\Gamma_1}{\Gamma_2}{\Gamma}$ and $\mrg{\Delta_1}{\Delta_2}{\Gamma_1}$ are valid, then there exists $\Delta$ such that $\mrg{\Delta_1}{\Gamma_2}{\Delta}$ and $\mrg{\Delta}{\Delta_2}{\Gamma}$ are valid.
\end{lemma}
\begin{proof}
  By induction on the derivation of $\mrg{\Gamma_1}{\Gamma_2}{\Gamma}$.
\end{proof}

\begin{lemma}
  For contexts $\Gamma_1, \Gamma_2, \Gamma, \Delta_1, \Delta_2$, if $\mrg{\Gamma_1}{\Gamma_2}{\Gamma}$ and $\mrg{\Delta_1}{\Delta_2}{\Gamma_1}$ are valid, then there exists $\Delta$ such that $\mrg{\Delta_2}{\Gamma_2}{\Delta}$ and $\mrg{\Delta_1}{\Delta}{\Gamma}$ are valid.
\end{lemma}
\begin{proof}
  By induction on the derivation of $\mrg{\Gamma_1}{\Gamma_2}{\Gamma}$.
\end{proof}

@@

\begin{lemma}\label{sub1conv}
  For terms $A, B, C$, if there are $A \prec B$ and $B \equiv C$, then there is $A \preceq C$.
\end{lemma}
\begin{proof}
  By \rname{$\prec$-$\preceq$} and the transitivity of equality $\equiv$.
\end{proof}

\begin{lemma}\label{convsub1}
  For terms $A, B, C$, if there are $A \equiv B$ and $B \prec C$, then there is $A \preceq C$.
\end{lemma}
\begin{proof}
  By \rname{$\prec$-$\preceq$} and the transitivity of equality $\equiv$.
\end{proof}

@@

\begin{lemma}\label{sub1trans}
  For terms $A, B, C, D$, if there are $A \prec B$, $B \equiv C$ and $C \prec D$, then there is $A \preceq D$.
\end{lemma}
\begin{proof}
  By induction on the derivation of $A \prec B$, definition of $\prec$ and Lemmas \ref{sub1sub}, \ref{sub1conv}, \ref{convsub1}.
\end{proof}

\begin{lemma}
  For terms $A, B, C$, if there are $A \preceq B$ and $B \preceq C$, then there is $A \preceq C$.
\end{lemma}
\begin{proof}
  By transitivity of $\equiv$, rule \rname{$\prec$-$\preceq$} and Lemma \ref{sub1trans}.
\end{proof}

\begin{lemma}
  For sorts $s, t$ and natural numbers $i, j$, if there is $s_i \preceq t_j$, then there are $s = t$ and $i \leq j$.
\end{lemma}
\begin{proof}
  By transitivity of $\equiv$ and Corollary \ref{sortinj}.
\end{proof}

\begin{lemma}
  For terms $A_1, A_2, B_1, B_2$ and sorts $s, t$, if there is $(x \stype{s} A_1) \rightarrow B_1 \preceq (x \stype{t} A_2) \rightarrow B_2$, then there are $A_1 \equiv A_2$, $B_1 \preceq B_2$ and $s = t$.
\end{lemma}
\begin{proof}
  By transitivity of $\equiv$ and Corollary \ref{arrowinj}.
\end{proof}

\begin{lemma}
  For terms $A_1, A_2, B_1, B_2$ and sorts $s, t$, if there is $(x \stype{s} A_1) \multimap B_1 \preceq (x \stype{t} A_2) \multimap B_2$, then there are $A_1 \equiv A_2$, $B_1 \preceq B_2$ and $s = t$.
\end{lemma}
\begin{proof}
  By transitivity of $\equiv$ and Corollary \ref{lolliinj}.
\end{proof}

@@

\begin{lemma}\label{uarrowinv}
  For any context $\Gamma$ and terms $A, B, s$, if there is $\Gamma \vdash (x \utype A) \rightarrow B : s$, then there exist sort $t$ and natural number $i$ such that $\Gamma \vdash A : U_i$ and $\Gamma, x \utype A \vdash B : t_i$.
\end{lemma}
\begin{proof}
  By induction on the derivation of $\Gamma \vdash (x \utype A) \rightarrow B : s$.
\end{proof}

\begin{lemma}\label{larrowinv}
  For any context $\Gamma$ and terms $A, B, s$, if there is $\Gamma \vdash (x \ltype A)\rightarrow B : s$, then there exist sort $t$ and natural number $i$ such that $\Gamma \vdash A : L_i$ and $\Gamma \vdash B : t_i$.
\end{lemma}
\begin{proof}
  By induction on the derivation of $\Gamma \vdash (x \ltype A) \rightarrow B : s$.
\end{proof}

\begin{lemma}\label{ulolliinv}
  For any context $\Gamma$ and terms $A, B, s$, if there is $\Gamma \vdash (x \utype A) \multimap B : s$, then there exist sort $t$ and natural number $i$ such that $\Gamma \vdash A : U_i$ and $\Gamma, x \utype A \vdash B : t_i$.
\end{lemma}
\begin{proof}
  By induction on the derivation of $\Gamma \vdash (x \utype A) \multimap B : s$.
\end{proof}

\begin{lemma}\label{llolliinv}
  For any context $\Gamma$ and terms $A, B, s$, if there is $\Gamma \vdash (x \ltype A)\multimap B : s$, then there exist sort $t$ and natural number $i$ such that $\Gamma \vdash A : L_i$ and $\Gamma \vdash B : t_i$.
\end{lemma}
\begin{proof}
  By induction on the derivation of $\Gamma \vdash (x \ltype A) \multimap B : s$.
\end{proof}

@@

\begin{lemma}\label{arrowlaminv}
  For any context $\Gamma$, terms $A, A', B, n$, sorts $s, s', t$ and natural number $i$, if there are $\overline{\Gamma} \vdash (x \stype{s'} A') \rightarrow B : t_i$ and $\Gamma \vdash \lambda x \stype{s} A . n : (x \stype{s'} A') \rightarrow B$, then there is $\Gamma, x \stype{s'} A' \vdash n : B$.
\end{lemma}
\begin{proof}
  Direct consequence of Lemmas \ref{uarrowinv}, \ref{larrowinv} and \ref{arrowlaminvx}.
\end{proof}

\begin{lemma}\label{lollilaminv}
  For any context $\Gamma$, terms $A, A', B, n$, sorts $s, s', t$ and natural number $i$, if there are $\overline{\Gamma} \vdash (x \stype{s'} A') \multimap B : t_i$ and $\Gamma \vdash \lambda x \stype{s} A . n : (x \stype{s'} A') \multimap B$, then there is $\Gamma, x \stype{s'} A' \vdash n : B$.
\end{lemma}
\begin{proof}
  Direct consequence of Lemmas \ref{ulolliinv}, \ref{llolliinv} and \ref{lollilaminvx}.
\end{proof}

@@

\begin{lemma}\label{agreerenpure}
  For contexts $\Gamma, \Gamma'$ and mapping $\xi$, if there are $agreeR\ \xi\ \Gamma\ \Gamma'$ and $\pure{\Gamma}$, then there is $\pure{\Gamma'}$.
\end{lemma}
\begin{proof}
  By induction on the derivation of $agreeR\ \xi\ \Gamma\ \Gamma'$.
\end{proof}

@@

\begin{lemma}\label{mergeagreereninv}
  For contexts $\Gamma, \Gamma', \Gamma_1, \Gamma_2$ and mapping $\xi$, if there are $agreeR\ \xi\ \Gamma\ \Gamma'$ and $\mrg{\Gamma_1}{\Gamma_2}{\Gamma}$, then there exist $\Gamma_1'$ and $\Gamma_2'$ such that there are $\mrg{\Gamma_1'}{\Gamma_2'}{\Gamma'}$, $agreeR\ \xi\ \Gamma_1\ \Gamma_1'$ and $agreeR\ \xi\ \Gamma_2\ \Gamma_2'$.
\end{lemma}
\begin{proof}
  By induction on the derivation of $agreeR\ \xi\ \Gamma\ \Gamma'$ and lemmas in Section \ref{agreeRprop}.
\end{proof}
